A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2023-05-22 JNABsub20230518/ComparingNLStools.tex
updates to bib

\texttt{NISTnls} (National Institutes for Standards and R port by Douglas Bates (2012)) -- This package provides R code and data for a set of
(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent real-world situations.

@@

\texttt{optimx} (John C. Nash and Varadhan (2011)) -- This provides
optimizers that can be applied to minimize a nonlinear function which could be
a nonlinear sum of
squares. NOT generally recommended if nonlinear
least squares programs can be easily used, but provides a check and alternative
solvers.

@@

The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth curve modeling
task that seems straightforward but is quite nasty. Its very succinct statement
provides the ``short reproducible example'' much requested on R mailing lists.
This is a real problem from a field researcher. The data and graph follow.

@@

\[   Asym =  b_1 = 100 * c_1 \]
\[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
\[ 1/scal  =  b_3 \]
To allow for simpler discussion, let us say that the parameters form a (named) vector
\(p\) and the model function is called \(model(p)\). The residuals can be written
either as \(r_1 = y - model(p)\) or \(r = model(p) - y\)
since their sum of squares has an identical value. The second form
avoids a potential sign error if we need to evaluate derivatives.

@@

The usual suggestion is that \(D\) be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of \(J' J\). In low precision, some elements of
\(J' J\) could underflow to zero (John C. Nash (1977)), and a linear combination of both
choices is an effective compromise. Various choices for
\(D\), as well as a possible line search along the
direction \(\delta\) rather than a unit step (Hartley (1961)),
give rise to several variant
algorithms. ``Marquardt's method'' is a family of methods. Fortunately, most choices
work well.

@@

@@

Additional termination tests can be used. \texttt{nlsr} has a \textbf{small sum of
squares} test (\textbf{smallsstest}) that compares the latest evaluated sum of squared
(weighted) residuals to \texttt{e4} times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is approximately 2.43e-55.
\%\% \^{} is picked up in rjtools check

@@

Tools that produce a class \texttt{nls} output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
\texttt{nlsr} computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian
are actually computed in the \texttt{print} and \texttt{summary} methods for the result.
These two approaches
lead to different consequences for performance and how features are
provided. \texttt{nlsr} has antecedents in the methods of John C. Nash (1979), where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus the code in \texttt{nlsr} is likely of value for workers to copy and modify
for customized tools.

@@

Gauss-Newton/Marquardt methods all need a Jacobian matrix at each iteration.
By default, \texttt{nlsr::nlxb()} will try to evaluate this using analytic
expressions using symbolic and automatic differentiation tools.
When using a formula specification of the model, \texttt{nls()}, \texttt{minpack.lm::nlsLM()}
and \texttt{gslnls::gsl\_nls()} use a finite difference approximation to compute the
Jacobian, though \texttt{gsl\_nls()} does have an option to attempt symbolic expressions.
Package \texttt{nlsr} provides, via appropriate calling syntax, four numeric approximation
options for the Jacobian, with a further control \texttt{ndstep} for the size of the
step used in the approximation.

@@

Users may also want to profit from the Jacobian code of \texttt{selfStart} models
but supply explicit starting values other than those suggested by \texttt{getInitial()}.
This does not appear to be possible with \texttt{nls()}. \texttt{nlsr::nlxb()} always requires
starting parameters, and can either use \texttt{getInitial()} to find them from the \texttt{selfStart}
model or provide explicit values, but the \texttt{selfStart} model is used in the formula
for the \texttt{nlxb()} call.

We are also surprised that the
analytic expressions for the Jacobian (``gradient'') in the \texttt{SSLogis} function
and others save quantities in ``hidden''
variables, i.e., with names preceded by ``.''.
These are then not displayed by the \texttt{ls()}
command, making them difficult to access by users who may wish to create
their own selfStart model via copy and edit.
Interactive tools, such as ``visual fitting'' (John C. Nash and Velleman (1996)) might be
worth considering as another way to find starting parameters, but we know of no
R capability of this type.

@@

\texttt{nls()} can be called without specifying the \texttt{data} argument. In this case, it will
search in the available environments (i.e., workspaces) for suitable data objects.
We do NOT like this approach, but it is ``the R way''. R allows users to leave many
objects in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to actively
suppress saving this workspace (\texttt{.RData}) on exit; otherwise, any such file in
the path will be loaded on startup. R users in our
acquaintance avoid saving the workspace because of lurking data and
functions that may cause unwanted results.

@@

\texttt{nls()} and other class \texttt{nls} tools accept an argument \texttt{subset}. This acts through
the mediation of \texttt{model.frame}, which not obvious in the
source code files \texttt{/src/library/stats/R/nls.R} and \texttt{/src/library/stats/src/nls.C}.
Having \texttt{subset} at the level of the call to a function like
\texttt{nls()} saves effort, but it does mean that the programmer of the
solver needs to be aware of the
origin (and value) of objects such as the data, residuals and Jacobian.
By preference,
we would implement subsetting by zero-value weights, with observation counts
(and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
we would extract a working dataframe from the relevant elements in the original.

@@

The function \texttt{resid()} (an alias for \texttt{residuals()}) gives WEIGHTED residuals,
as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}
allows us to compute residuals for particular coefficient sets.
We have had to extract from the base R code and include it via a code chunk
(not echoed here for space) because it is not exported to the working namespace,
We could also explicitly \texttt{source()} this code.

@@

\begin{verbatim}
## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
attr(rn0,"gradient")<-NULL; rn0
\end{verbatim}

@@

\begin{verbatim}
nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
rn<-nmod$resid()
attr(rn,"gradient")<-NULL; rn
\end{verbatim}

@@

Note that the sum of squares has been reduced from 61216 to 1.6211, but
unless \texttt{trace} is invoked, the user will not get any information about this.
This almost trivial change to the \texttt{nls()} function could be
useful to R users.

@@

The variable projection method (Golub and Pereyra (1973), O'Leary and Rust (2013)) is
generally much more effective than general approaches in finding good solutions
to nonlinear least squares
problems when some of the parameters appear linearly. In our logistic examples,
the asymptote parameters are an illustration. However, identifying which parameters
are linear and communicating this information to estimating functions
is not a trivial task. \texttt{nls()} has an option \texttt{algorithm="plinear"} that allows some partially linear
models to be solved. The other tools, as far as we are aware, do not offer any
such capability. The \texttt{nlstac} package uses a different algorithm for similar goals.

@@

Some models have several common parameters and
others that are tied to particular cases.
The \textbf{man} file for \texttt{nls()} includes an example of a situation in which
parameters are indexed but which uses the ``plinear'' option as an added complication.
Running this example reveals that the answers for the parameters are NOT indexed
as in a vector.
That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}} but \texttt{a1,\ a2,\ a3}.
This is no doubt because programming for indexed parameters is challenging.
We note that there are capabilities in packages for mixed-effects modeling
such as \texttt{nlme} (Pinheiro et al. (2013)), \texttt{bbmle} (Bolker and Team (2013)), and \texttt{lme4} (p-lme4) for
estimating models of such type.

@@

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{references-1}{%
\section*{References}\label{references-1}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-R-onls-manual}{}}%
Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.

\leavevmode\vadjust pre{\hypertarget{ref-bateswatts}{}}%
Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.

\leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.''} \emph{Technometrics} 23 (2): 179--83.

\leavevmode\vadjust pre{\hypertarget{ref-p-nlstools}{}}%
Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools: Tools for Nonlinear Regression Diagnostics}.

\leavevmode\vadjust pre{\hypertarget{ref-nlsMicrobio-manual}{}}%
---------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.

\leavevmode\vadjust pre{\hypertarget{ref-p-bbmle}{}}%
Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for General Maximum Likelihood Estimation}. \url{http://CRAN.R-project.org/package=bbmle}.

\leavevmode\vadjust pre{\hypertarget{ref-gslnls23}{}}%
Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}. \url{https://CRAN.R-project.org/package=gslnls}.

\leavevmode\vadjust pre{\hypertarget{ref-nlshelper-manual}{}}%
Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for Non-Linear Regression}. \url{https://CRAN.R-project.org/package=nlshelper}.

\leavevmode\vadjust pre{\hypertarget{ref-minpacklm12}{}}%
Elzhov, Timur V., Katharine M. Mullen, Andrej-Nikolai Spiess, and Ben Bolker. 2012. \emph{Minpack.lm: R Interface to the Levenberg-Marquardt Nonlinear Least-Squares Algorithm Found in MINPACK, Plus Support for Bounds}. R Project for Statistical Computing. \url{http://CRAN.R-project.org/package=minpack.lm}.

\leavevmode\vadjust pre{\hypertarget{ref-GSL-manual}{}}%
Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for GSL Version 1.12 (3. Ed.).}

\leavevmode\vadjust pre{\hypertarget{ref-Bioconductor}{}}%
Gentleman, Robert C., Vincent J. Carey, Douglas M. Bates, Ben Bolstad, Marcel Dettling, Sandrine Dudoit, Byron Ellis, et al. 2004. {``Bioconductor: Open Software Development for Computational Biology and Bioinformatics.''} \emph{{Genome Biology}} 5 (R80). \url{https://doi.org/10.1186/gb-2004-5-10-r80}.

\leavevmode\vadjust pre{\hypertarget{ref-Golub1973}{}}%
Golub, G. H., and V. Pereyra. 1973. {``The Differentiation of Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.

\leavevmode\vadjust pre{\hypertarget{ref-nls2-22}{}}%
Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute Force}. \url{https://CRAN.R-project.org/package=nls2}.

\leavevmode\vadjust pre{\hypertarget{ref-Hartley1961}{}}%
Hartley, H. O. 1961. {``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.''} \emph{Technometrics} 3: 269-\/-280.

\leavevmode\vadjust pre{\hypertarget{ref-Huet2004}{}}%
Huet, S., A. Bouvier, M.-A. Poursat, and E. Jolivet. 2004. \emph{Statistical Tools for Nonlinear Regression: A Practical Guide with {S-PLUS} Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.

\leavevmode\vadjust pre{\hypertarget{ref-nlsr2023manual}{}}%
John C Nash, and Duncan Murdoch. 2023. \emph{{nlsr: Functions for Nonlinear Least Squares Solutions}}.

\leavevmode\vadjust pre{\hypertarget{ref-tvrdik2007}{}}%
Josef Tvrdík and Ivan Křivý and Ladislav Mišík. 2007. {``Adaptive Population-Based Search: Application to Estimation of Nonlinear Regression Parameters.''} \emph{{Computational Statistics \& Data Analysis}} 52 (2): 713--24. \url{https://doi.org/10.1016/j.csda.2006.10.014}.

\leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied Mathematics} 2: 164-\/-168.

\leavevmode\vadjust pre{\hypertarget{ref-Marquardt1963}{}}%
Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.

\leavevmode\vadjust pre{\hypertarget{ref-MiguezNLRAA2021}{}}%
Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for Agricultural Applications}}. \url{https://CRAN.R-project.org/package=nlraa}.

\leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.''} \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.

\leavevmode\vadjust pre{\hypertarget{ref-cnm79}{}}%
---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.

\leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
---------. 2022. {``{Function minimization and nonlinear least squares in R}.''} \emph{WIREs Computational Statistics}, no. e1580. https://doi.org/\url{https://doi.org/10.1002/wics.1580}.

\leavevmode\vadjust pre{\hypertarget{ref-PkgFromRbase22}{}}%
Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package from Base {R} Files.''} \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.

\leavevmode\vadjust pre{\hypertarget{ref-p-optimx}{}}%
Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A Replacement and Extension of the {optim()} Function}. Nash Information Services Inc.; Johns Hopkins University.

\leavevmode\vadjust pre{\hypertarget{ref-nash1996nonlinear}{}}%
Nash, John C., and Paul Velleman. 1996. {``Nonlinear Estimation Combining Visual Fitting with Optimization Methods.''} In \emph{Proceedings of the Section on Physical and Engineering Sciences of the American Statistical Association}, 256--61. American Statistical Association.

\leavevmode\vadjust pre{\hypertarget{ref-NISTnls-manual}{}}%
National Institutes for Standards, original from, and Technology http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by Douglas Bates. 2012. \emph{NISTnls: Nonlinear Least Squares Examples from NIST}. \url{https://CRAN.R-project.org/package=NISTnls}.

\leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for Nonlinear Least Squares Problems.''} \emph{Computational Optimization and Applications} 54 (3): 579--93.

\leavevmode\vadjust pre{\hypertarget{ref-nlsmultstart}{}}%
Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart: Robust Non-Linear Regression Using AIC Scores}. \url{https://CRAN.R-project.org/package=nls.multstart}.

\leavevmode\vadjust pre{\hypertarget{ref-p-nlme}{}}%
Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.

\leavevmode\vadjust pre{\hypertarget{ref-m-nlsmsn}{}}%
Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn: Fitting Nonlinear Models with Scale Mixture of Skew-Normal Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.

\leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified Practical Approach}. New York; Basel: Marcel Dekker Inc.

>>>>

\texttt{NISTnls} ((\emph{NISTnls: Nonlinear Least Squares Examples from Nist} 2012)) -- This package provides R code and data for a set of
(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent real-world situations.

@@

\texttt{optimx} (Nash and Varadhan (2011)) -- This provides
optimizers that can be applied to minimize a nonlinear function which could be
a nonlinear sum of
squares. NOT generally recommended if nonlinear
least squares programs can be easily used, but provides a check and alternative
solvers.

@@

The Hobbs weed infestation problem (Nash (1979, 120)) is a growth curve modeling
task that seems straightforward but is quite nasty. Its very succinct statement
provides the ``short reproducible example'' much requested on R mailing lists.
This is a real problem from a field researcher. The data and graph follow.

@@

\[   Asym =  b_1 = 100 * c_1 \]
\[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
\[ 1/scal  =  b_3 = 0.1*c3 \]
To allow for simpler discussion, let us say that the parameters form a (named) vector
\(p\) and the model function is called \(model(p)\). The residuals can be written
either as \(r_1 = y - model(p)\) or \(r = model(p) - y\)
since their sum of squares has an identical value. The second form
avoids a potential sign error if we need to evaluate derivatives.

@@

The usual suggestion is that \(D\) be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of \(J' J\). In low precision, some elements of
\(J' J\) could underflow to zero (Nash (1977)), and a linear combination of both
choices is an effective compromise. Various choices for
\(D\), as well as a possible line search along the
direction \(\delta\) rather than a unit step (Hartley (1961)),
give rise to several variant
algorithms. ``Marquardt's method'' is a family of methods. Fortunately, most choices
work well.

@@

\begin{verbatim}
#> unlx1:
\end{verbatim}

@@

Additional termination tests can be used. \texttt{nlsr} has a \textbf{small sum of
squares} test (\textbf{smallsstest}) that compares the latest evaluated sum of squared
(weighted) residuals to \texttt{e4} times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is approximately 2.43e-55.

@@

Tools that produce a class \texttt{nls} output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
\texttt{nlsr} computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian
are actually computed in the \texttt{print} and \texttt{summary} methods for the result.
These two approaches
lead to different consequences for performance and how features are
provided. \texttt{nlsr} has antecedents in the methods of Nash (1979), where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus the code in \texttt{nlsr} is likely of value for workers to copy and modify
for customized tools.

@@

Gauss-Newton/Marquardt methods all need a Jacobian matrix at each iteration.
By default, \texttt{nlsr::nlxb()} will try to evaluate this using analytic
expressions using symbolic and automatic differentiation tools.
When using a formula specification of the model, \texttt{nls()}, \texttt{minpack.lm::nlsLM()}
and \texttt{gslnls::gsl\_nls()} use a finite difference approximation to compute the
Jacobian, though \texttt{gsl\_nls()} does have an option to attempt symbolic expressions.
Package \texttt{nlsr} provides, via appropriate calling syntax, four numeric approximation
options for the Jacobian, with a further control \texttt{ndstep} for the size of the
step used in the approximation. These options allow programming choices to be
examined. Users can largely ignore them.

@@

Users may also want to profit from the Jacobian code of \texttt{selfStart} models
but supply explicit starting values other than those suggested by \texttt{getInitial()}.
This does not appear to be possible with \texttt{nls()}. \texttt{nlsr::nlxb()} always requires
starting parameters, and can either use \texttt{getInitial()} to find them from the \texttt{selfStart}
model or provide explicit values, but the formula used in the call to \texttt{nlxb()}
still involves the \texttt{selfStart} function.

We are also surprised that the
analytic expressions for the Jacobian (``gradient'') in the \texttt{SSLogis} function
and others save quantities in ``hidden''
variables, i.e., with names preceded by ``.''.
These are then not displayed by the \texttt{ls()}
command, making them difficult to access by users who may wish to create
their own selfStart model via copy and edit.
Interactive tools, such as ``visual fitting'' (Nash and Velleman (1996)) might be
worth considering as another way to find starting parameters, but we know of no
R capability of this type.

@@

\texttt{nls()} can be called without specifying the \texttt{data} argument. In this case, it will
search in the available environments (i.e., workspaces) for suitable data objects.
We do NOT like this approach, but it is ``the R way''. R allows users to leave many
objects in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to actively
suppress saving this workspace (\texttt{.RData}) on exit; otherwise any such file in
the path will be loaded on startup. R users in our
acquaintance avoid saving the workspace because of lurking data and
functions that may cause unwanted results.

@@

\texttt{nls()} and other class \texttt{nls} tools accept an argument \texttt{subset}. This acts through
the mediation of \texttt{model.frame}, which is not obvious in the
source code files
\texttt{/src/library/stats/R/nls.R} and
\texttt{/src/library/stats/src/nls.C}.
Having \texttt{subset} at the level of the call to a function like
\texttt{nls()} saves effort, but it does mean that the programmer of the
solver needs to be aware of the
origin (and value) of objects such as the data, residuals and Jacobian.
By preference,
we would implement subsetting by zero-value weights, with observation counts
(and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
we would extract a working dataframe from the relevant elements in the original.

@@

The function \texttt{resid()} (an alias for \texttt{residuals()}) gives WEIGHTED residuals,
as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}
allows us to compute residuals for particular coefficient sets.
We have had to extract \texttt{nlsModel()} from the base R code and include it via a code chunk
(not echoed here for space) because it is not exported to the working namespace,
We could also explicitly \texttt{source()} this code.

@@

\begin{verbatim}
## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
attr(rn0,"gradient")<-NULL; rn0 # weighted residuals at starting coefficients
\end{verbatim}

@@

\begin{verbatim}
nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
rn<-nmod$resid()
attr(rn,"gradient")<-NULL; rn # same as rusewt
\end{verbatim}

@@

Note that the sum of squares has been reduced from 61216 to 1.6211, but
unless \texttt{trace} is invoked, the user will not get any information about this.
Changing this in the \texttt{nls()} function could be useful to R users, and would
be almost trivial.

@@

The variable projection method (Golub and Pereyra (1973), O'Leary and Rust (2013)) is
usually much more effective than general approaches in finding good solutions
to nonlinear least squares
problems when some of the parameters appear linearly. In our logistic examples,
the asymptote parameters are an illustration. However, identifying which parameters
are linear and communicating this information to estimating functions
is not a trivial task. \texttt{nls()} has an option \texttt{algorithm="plinear"} that allows some partially linear
models to be solved. The other tools, as far as we are aware, do not offer any
such capability. The \texttt{nlstac} package uses a different algorithm for similar goals.

@@

Some models have several common parameters and
others that are tied to particular cases.
The \textbf{man} file for \texttt{nls()} includes an example of a situation in which
parameters are indexed but which uses the ``plinear'' option as an added complication.
Running this example reveals that the answers for the parameters are NOT indexed
as in a vector.
That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}} but \texttt{a1,\ a2,\ a3}.
This is no doubt because programming for indexed parameters is challenging.
We note that there are capabilities in packages for mixed-effects modeling
such as \texttt{nlme} (Pinheiro et al. (2013)), \texttt{bbmle} (Bolker and Team (2013)), and \texttt{lme4} (Bates et al. (2015)) for
estimating models of such type.

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-R-onls-manual}{}%
Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.

\leavevmode\hypertarget{ref-bateswatts}{}%
Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.

\leavevmode\hypertarget{ref-p-lme4}{}%
Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015. ``Fitting Linear Mixed-Effects Models Using lme4.'' \emph{Journal of Statistical Software} 67 (1): 1--48. \url{https://doi.org/10.18637/jss.v067.i01}.

\leavevmode\hypertarget{ref-BatesWatts81}{}%
Bates, Douglas M., and Watts, Donald G. 1981. ``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.'' \emph{Technometrics} 23 (2): 179--83.

\leavevmode\hypertarget{ref-p-nlstools}{}%
Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools: Tools for Nonlinear Regression Diagnostics}.

\leavevmode\hypertarget{ref-nlsMicrobio-manual}{}%
---------. 2014. \emph{NlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.

\leavevmode\hypertarget{ref-p-bbmle}{}%
Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for General Maximum Likelihood Estimation}. \url{http://CRAN.R-project.org/package=bbmle}.

\leavevmode\hypertarget{ref-gslnls23}{}%
Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}. \url{https://CRAN.R-project.org/package=gslnls}.

\leavevmode\hypertarget{ref-nlshelper-manual}{}%
Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for Non-Linear Regression}. \url{https://CRAN.R-project.org/package=nlshelper}.

\leavevmode\hypertarget{ref-minpacklm12}{}%
Elzhov, Timur V., Katharine M. Mullen, Andrej-Nikolai Spiess, and Ben Bolker. 2012. \emph{Minpack.lm: R Interface to the Levenberg-Marquardt Nonlinear Least-Squares Algorithm Found in Minpack, Plus Support for Bounds}. R Project for Statistical Computing. \url{http://CRAN.R-project.org/package=minpack.lm}.

\leavevmode\hypertarget{ref-GSL-manual}{}%
Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for Gsl Version 1.12 (3. Ed.).}

\leavevmode\hypertarget{ref-Bioconductor}{}%
Gentleman, Robert C., Vincent J. Carey, Douglas M. Bates, Ben Bolstad, Marcel Dettling, Sandrine Dudoit, Byron Ellis, et al. 2004. ``Bioconductor: Open Software Development for Computational Biology and Bioinformatics.'' \emph{Genome Biology} 5 (R80). \url{https://doi.org/10.1186/gb-2004-5-10-r80}.

\leavevmode\hypertarget{ref-Golub1973}{}%
Golub, G. H., and V. Pereyra. 1973. ``The Differentiation of Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables Separate.'' \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.

\leavevmode\hypertarget{ref-nls2-22}{}%
Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute Force}. \url{https://CRAN.R-project.org/package=nls2}.

\leavevmode\hypertarget{ref-Hartley1961}{}%
Hartley, H. O. 1961. ``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.'' \emph{Technometrics} 3: 269--80.

\leavevmode\hypertarget{ref-Huet2004}{}%
Huet, S., A. Bouvier, M.-A. Poursat, and E. Jolivet. 2004. \emph{Statistical Tools for Nonlinear Regression: A Practical Guide with S-PLUS Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.

\leavevmode\hypertarget{ref-nlsr2023manual}{}%
John C Nash, and Duncan Murdoch. 2023. \emph{nlsr: Functions for Nonlinear Least Squares Solutions}.

\leavevmode\hypertarget{ref-tvrdik2007}{}%
Josef Tvrdík and Ivan Křivý and Ladislav Mišík. 2007. ``Adaptive Population-Based Search: Application to Estimation of Nonlinear Regression Parameters.'' \emph{Computational Statistics \& Data Analysis} 52 (2): 713--24. \url{https://doi.org/10.1016/j.csda.2006.10.014}.

\leavevmode\hypertarget{ref-Levenberg1944}{}%
Levenberg, Kenneth. 1944. ``A Method for the Solution of Certain Non-Linear Problems in Least Squares.'' \emph{Quarterly of Applied Mathematics} 2: 164--68.

\leavevmode\hypertarget{ref-Marquardt1963}{}%
Marquardt, Donald W. 1963. ``An Algorithm for Least-Squares Estimation of Nonlinear Parameters.'' \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.

\leavevmode\hypertarget{ref-MiguezNLRAA2021}{}%
Miguez, Fernando. 2021. \emph{nlraa: Nonlinear Regression for Agricultural Applications}. \url{https://CRAN.R-project.org/package=nlraa}.

\leavevmode\hypertarget{ref-jn77ima}{}%
Nash, John C. 1977. ``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.'' \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.

\leavevmode\hypertarget{ref-cnm79}{}%
---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.

\leavevmode\hypertarget{ref-NashWires22}{}%
---------. 2022. ``Function minimization and nonlinear least squares in R.'' \emph{WIREs Computational Statistics}, no. e1580. \url{https://doi.org/https://doi.org/10.1002/wics.1580}.

\leavevmode\hypertarget{ref-PkgFromRbase22}{}%
Nash, John C., and Arkajyoti Bhattacharjee. 2022. ``Making a Package from Base R Files.'' \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.

\leavevmode\hypertarget{ref-p-optimx}{}%
Nash, John C, and Ravi Varadhan. 2011. \emph{\textup{Optimx}: A Replacement and Extension of the optim() Function}. Nash Information Services Inc.; Johns Hopkins University.

\leavevmode\hypertarget{ref-nash1996nonlinear}{}%
Nash, John C., and Paul Velleman. 1996. ``Nonlinear Estimation Combining Visual Fitting with Optimization Methods.'' In \emph{Proceedings of the Section on Physical and Engineering Sciences of the American Statistical Association}, 256--61. American Statistical Association.

\leavevmode\hypertarget{ref-NISTnls-manual}{}%
\emph{NISTnls: Nonlinear Least Squares Examples from Nist}. 2012. \url{https://CRAN.R-project.org/package=NISTnls}.

\leavevmode\hypertarget{ref-OlearyRust13}{}%
O'Leary, Dianne P., and Bert W. Rust. 2013. ``Variable Projection for Nonlinear Least Squares Problems.'' \emph{Computational Optimization and Applications} 54 (3): 579--93.

\leavevmode\hypertarget{ref-nlsmultstart}{}%
Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart: Robust Non-Linear Regression Using Aic Scores}. \url{https://CRAN.R-project.org/package=nls.multstart}.

\leavevmode\hypertarget{ref-p-nlme}{}%
Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.

\leavevmode\hypertarget{ref-m-nlsmsn}{}%
Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn: Fitting Nonlinear Models with Scale Mixture of Skew-Normal Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.

\leavevmode\hypertarget{ref-Ratkowsky1983}{}%
Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified Practical Approach}. New York; Basel: Marcel Dekker Inc.

\leavevmode\hypertarget{ref-m-nlstac}{}%
Rodriguez-Arias, Mariano, Juan Antonio Fernandez, Javier Cabello, and Rafael Benitez. 2020. \emph{Nlstac: An R Package for Fitting Separable Nonlinear Models}. \url{https://CRAN.R-project.org/package=nlstac}.

\leavevmode\hypertarget{ref-nlsic-manual}{}%
Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.

\leavevmode\hypertarget{ref-Torvisco2018}{}%
Torvisco, Juan Antonio Fernandez, Mariano Rodrıguez-Arias Fernandez, and Javier Cabello Sanchez. 2018. ``A New Algorithm to Fit Exponential Decays Without Initial Guess.'' Faculty of Sciences and Mathematics, University of Niˇs, Serbia. \url{https://doi.org/doi.org/10.2298/FIL1812233T}.
